sdmTMB statistical model description

### 2021-11-30 doc/appendix-model.Rmd
Add to model appendix

# Basic model structure

@@

We now split this model apart to describe the various parts in more detail

## Fixed- or main-effect portion:

@@

@@

If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept.

@@

# Threshold models

## Linear breakpoint threshold models

## Logistic threshold models

$$
s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
     / \left(s95 - s50 \right) } \right] }^{-1},
$$
where $\psi$ is a scaling parameter (controlling the height of the y-axis for
the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
a parameter controlling the point at which the function reaches 50% of the
maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
constrained to be larger than $s50$.

@@

@@

The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as 
$$
\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1} (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right),$$
<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents a scaling parameter that controls the decorrelation distance (REF).
The parameter $\nu$ controls the smoothness of the covariance function.
We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].

>>>>

# Combined model structure

@@

We now split this model apart to describe the various parts in more detail.

## Main effects:

@@

### Smoothers

### Linear breakpoint threshold models

### Logistic threshold models

$$
s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
     / \left(s95 - s50 \right) } \right] }^{-1},
$$

where $\psi$ is a scaling parameter (controlling the height of the y-axis for
the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
a parameter controlling the point at which the function reaches 50% of the
maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
constrained to be larger than $s50$.

@@

If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).

@@

@@

# Random field details

@@

The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as

$$
\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1} (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right),$$
<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->

where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents a scaling parameter that controls the decorrelation distance (REF).
The parameter $\nu$ controls the smoothness of the covariance function.
We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].

## Projection 'A' matrix

## Anisoptropy

## Barrier mesh

# Optimization details

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].

TODO two-stage fitting

TODO extra loops and Newton loops

# Assessing convergence
